Learning to Optimize Real-World Programs

### 2021-02-04 paper.tex
4:30

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

@@

@@

We perform a two-step process for learning program optimization.
In this section, we first describe our underlying model, then describe the first step of pre-training the model using imitation learning, and finally describe the second step of using a hill-climbing algorithm that directly minimizes the cost of model-optimized programs \gn{Does it actually minimize the expected cost (in a mathematical sense)? I don't think so.}\pcyin{Done: deleted ``expected''}.
%\pcyin{We need to change the term ``fine-tuning''. Can we say ``RL'' here? Perhaps not because it's not exactly RL?}
%using the outputs of the neural model itself. 
%For the first phase, the dataset of 1.77 million functions mined from open source projects was used. In the second phase we fine-tune the pre-trained model using an iterative-hill climbing algorithm.

\subsection{Underlying Model}

@@

Our model $f_{\theta}$ is based on a standard Transformer-based encoder \citep{vaswani2017attention}, but with a LSTM-based decoder \citep{hochreiter1997long,chen-etal-2018-best} instead.
This was mainly done to speed-up prediction within the iterative training loop described in \S\ref{sec:hillclimbing}, as transformers suffer from quadratic decoding time in sequence length \citep{chen-etal-2018-best}. We defer more implementation-specific details for our experiments to Section \ref{sec:experiments}.

@@

\subsection{Pre-training}

In the pre-training step, we train a neural model learn common optimizations performed by an optimizing compiler (e.g., \texttt{GCC}) using maximum likelihood estimation. \alex{I think we could remove this section, and mention we do pre-training before hill-climbing in section 6}

@@

\subsection{Iterative Hill-Climbing}

\pcyin{When explaining a model/algorithm, it's important to first briefly describe the intuition behind each component/sub-section at the very beginning. Can we say something like the follows here: With the model pre-trained on compiler-optimized examples, we then aim to learn more efficient optimization strategies by directly minimizing the expected cost of model-optimized programs.}
Policy gradient algorithms seem like a natural fit for learning how to optimize programs; however, methods based on maximum-likelihood estimation, or imitation learning, provide advantages in terms of stability. Training large neural networks with policy gradient algorithms can be difficult to tune. These difficulties are often attributed to the gradient's high variance which is exacerbated by other issues in program synthesis: a general sparsity of rewards, credit assignment over long sequences, and brittleness of program semantics.

@@

If an expert policy or expert examples are available, imitation learning can be desirable, because of its stability. Impressive results have been achieved in the game of computer Go and other games by minimizing the cross-entropy between a policy and an expert-like Monte Carlo tree search policy \cite{silver2017mastering}. In program synthesis from I/O example tasks, supervised learning on example programs has been shown to be competitive, and at times even superior to RL \cite{bunel2018leveraging}. The effectiveness of supervised, or imitation learning, on program synthesis tasks is even more surprising, given the programs used as targets were generated from random generation from a grammar. Given that real-world programs demonstrate a degree of naturalness, and thus, a higher degree of predictability and repetition \cite{hindle2012naturalness}, one may expect imitation learning to generally perform even better in practice on general program synthesis tasks. \alex{I feel these 2 sentences are good points but could be a little too much detail.}\pcyin{I agree. We could try shortening this paragraph once Section 6 is nearly done.}

@@

Inspired by the success of imitation learning in program synthesis, in this paper we propose an intuitive hill-climbing algorithm, a special case of imitation learning, for learning program optimizations. 
%We use an intuitive iterative hill-climbing algorithm algorithm for learning program optimizations. 
%It consists of a two-step process in which an exploration batch $B_{\mathrm{ex}}$ is sampled from the model. 
It consists of a two-step process. 
First, an exploration batch $B_{\mathrm{ex}}$ is sampled from the dataset $D$ initialized with mined programs $F^i_u$ and their compiler-optimized outputs $F^i_o$. 
For each input program $F^i_u$ in $B_{\mathrm{ex}}$, we sample a model-predicted optimization $\hat{F}^u_o$, and run $\hat{F}^u_o$ on the test the I/O test suite to compute the cost function $\mathcal{C}(\cdot)$.
%Each output from the batch is executed on the testcases to test equivalence and its performance characteristics are measured and profiled. 
If any of the new samples are both functionally equivalent by our distance function $\mathcal{D}(\cdot)$ and also more optimal compared to the compiler-optimized targets in the original dataset, the compiler-optimized target in the dataset is then replaced with the model's newly-discovered optimal rewrite. 
\pcyin{For each $F^i_u$, are multiple samples of $\hat{F}^i_o$ generated? If that's the case we need to modify Algo 1 to clarify this.}
After the exploration step is taken, a separate training batch $B_{\mathrm{tr}}$ is sampled for maximum-likelihood training from the dataset which may now contain model-optimized targets. \pcyin{The algorithm sketch here is good. We could improve the narrative by connecting each chunks of sentences to line numbers in Algo 1. See \url{https://www.overleaf.com/read/qmxxdyrnqjft}(lines 139 to 164) for an example.}

@@

We report our performance on a randomly-sampled subset of the training set as well as the entire test set. Our intent in doing so is to evaluate the learned model in two task-settings: the first being an \textit{inductive} task where we are interested in the model generated and its ability to generalize to unseen examples in the test set. The second task is a \textit{transductive} task where we are not interested in the model per-se, rather, we are interested in the capacity of the model and the training procedure to transduce unoptimized programs to optimized programs. In both tasks we generate samples using a beam-size of 10, executing all on the test suite $\{IO\}_{k}^K$ to calculate both $\mathcal{C}$  and $\mathcal{D}$ and we optionally use the SMT-based verifier to also calculate $D$. \footnote{Another way to perform transductive evaluation is by reporting statistcs on the best outputs generated during training. We felt beam-search would provide a more interesting comparison.}. Then, of these ten samples, we report results on the best-performing correct sample (if one was available).

@@

We subsequently performed a manual evaluation of the outputs. We randomly sampled 29 of the programs that beat the \texttt{GCC -Og} benchmark in the hold-out set. Upon close analysis, we found 8 of these 29 programs (or 27.59\%) were semantics preserving and correct. We witnessed two large groups of errors: the first was that the model would often remove conditional branches from programs that were unlikely to be triggered through random testing. 
%
%We witnessed that the model often %removed entire branches from programs. 
%\pcyin{Sorry I don't quite get the %following two sentences.}
Take an example in which we have a program in which we would like to trigger behavior if the variable $x$ is equal to the constant 0 (i.e. $\textbf{if} \;\; \mathbf{x\;==\;0:\; \textsc{do}...}$). If $x$ is represented by a 64-bit integer in a cpu register, and if we use a uniform random distribution to generate values for x in testcases,  likelihood of sampling the exact constant 0 is $(2^{-64})$, and thus very unlikely. As a result, randomly-generated testcases can be highly insufficient for measuring program correctness.

@@

% Given that randomly generated testcases were insufficient to reveal many spurious optimizations, we then evaluated our model outputs for correctness using the  bounded model checker for x86-64 assembly code available in \textsc{STOKE} which makes use of the Z3 SMT solver \cite{de2008z3, churchill2017sound}. 
% In theory, the model checker should be able to accurately check for correctness as long as the number of loop iterations stays below the configured maximum bound of 4, and the process does not timeout by taking over 180 seconds.
% We report these results in Table \ref{tab:z3_results}.

When evaluating with SMT solvers, the proportion of programs that outperform the -Og baseline drops dramatically. Nevertheless, the fine-tuning procedure still allows for a marginal improvement over the pre-trained baseline. Moreover, we see the proportion of programs for the fine-tuned model that beat the GCC -O0 baseline drops below the pre-trained model. These programs that were unable to beat the GCC baseline were in-fact declared by the SMT solver to not be logically equivalent: we hypothesize that the fine-tuned model over-fit on spurious programs through the hill-climbing process. As a result, it may have learned some new valid optimizations; yet it evidently lost its ability to more-consistently perform less-aggressive, semantics-preserving optimizations.

@@

Given that fine-tuning with testcases to measure equivalence seemed to bias the model to learn spurious optimizations, we additionally fine-tuned the model again using both testcases as well as the SMT solver to determine equivalence. We then evaluated this model only using both testcases as well as the solver. Our results are located in Table \ref{tab:z3_inner_loop_results}.

Utilizing the solver in the inner-loop of training did not cause a significant degradation in performance in the task of outperforming the -O0 baseline the way training on testcases only did, and on both metrics we see the fine-tuning process enabled the model to outperform the pre-trained baseline model. For thoroughness, we sampled \alex{XX} programs that outperformed the GCC -Og benchmark and manually evaluated their correcness and saw \alex{XX.XX\%, i.e. very few} were incorrect. A full breakdown of the manual analysis results are located in \ref{tab:manual}.

% \subsection{Manual Analysis}

% A different pattern emerged when examining programs evaluated with the theorem prover: in some cases the theorem prover had difficulty with more complex programs that performed memory references. Another pattern emerged in which it seems our heuristic greedy algorithm for selecting areas of hardware to measure for equivalence may have been too-permissive in allowing important areas of the hardware state to not count for equivalence checking. This may have originated, because certain pathological testcases may have caused the programs to differ in parts of the hardware state that one would expect to be equivalent.

\begin{table}[t]
\caption{Evaluation results for model fine-tuned on testcases-only and evaluated on testcases-only. The table report the proportion of programs sampled outperforming GCC settings. The inductive task is denoted by \textit{ind.} and the transductive task is denoted by \textit{trans.}}
\label{tab:random_generation_results}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcc}
\toprule
Task & Outperf. -O0 & Outperf. -Og \\
\midrule
Pre-Trained ind.    & 90.33\% & 5.67\% \\
Hill-Climbing ind.  & 87.56\% & 22.22\% \\
Pre-Trained trans.   & 89.43\% & 6.04\% \\
Hill-Climbing trans.  & 93.14\% & 24.27\% \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\begin{table}[t]
\caption{Evaluation results for model fine-tuned on testcases-only and evaluated with testcases and the SMT solver for correcteness. The table report the proportion of programs sampled outperforming GCC settings. The inductive task is denoted by \textit{ind.} and the transductive task is denoted by \textit{trans.}}
\label{tab:z3_results}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcc}
\toprule
Task & Outperf. -O0 & Outperf. -Og \\
\midrule
Pre-Trained ind.    & 86.22\% & 1.11 \% \\
Hill-Climbing ind.  & 65.33\% & 5.22\% \\
Pre-Trained trans.   & 85.93\% & 2.21\% \\
Hill-Climbing trans.  & 74.69\% & 5.81\% \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\begin{table}[t]
\caption{Evaluation results for model fine-tuned and evaluated with both testcases and the SMT solver for determining correctness. The table report the proportion of programs sampled outperforming GCC settings. The inductive task is denoted by \textit{ind.} and the transductive task is denoted by \textit{trans.}}
\label{tab:z3_inner_loop_results}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcc}
\toprule
Task & Outperf. -O0 & Outperf. -Og \\
\midrule
Pre-Trained ind.    & --.-\% & --.--\% \\
Hill-Climbing ind.  & --.--\% & --.--\% \\
Pre-Trained trans.   & --.--\% & --.--\% \\
Hill-Climbing trans.  & --.--\% & --.--\% \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\begin{table}[t]
\caption{This table reports the proportion of programs that manual evaluation determined correct. This was performed only for the examples within the hold-out test set for the inductive task and the models trained with the iterative hill-climbing algorithm and for the programs that outperformed GCC -Og.  \textbf{TC} corresponds to experiments where correctness was determined only by passing all testcases; whereas, \textbf{SMT} corresponds to experiments where correctness was determined by passing all testcases \textit{and} verifying via the SMT solver}
\label{tab:manual}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lc}
\toprule
Task & Proportion Verified \\
\midrule
TC Train/TC Eval  & 22.72\%\\
TC Train/Z3 Eval   & 91.30\%\\
Z3 Train/Z3 Eval  & --.--\%\\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

@@

@@

Most x86-64 instructions allow operands to be stored in any register.  A notable exception is the \es{decorate me} \texttt{mulq} instruction, which multiplies two 64-bit operands and stores the 128-bit result without truncation.
%
Though this instruction multiplies two numbers, it only takes one operand as an argument.  By definition, \texttt{mulq op} computes the multiplication of \texttt{\%rax} and \texttt{op}, and then stores the upper 64-bits of the result in \%rdx and the lower bits in \%rax.  The registers \%rax and \%rdx are fixed and cannot be changed.
%
\ed{@pcyin Can we put the two listings side by side?  We can truncate the long constant, it's not important.  It would also be nice if we could highlight the changes.}
\begin{figure}
% -Og compiler code
    \begin{lstlisting}
movq %rdi, %rdx
shrq $0x3, %rdx
movq $0x20c49ba5e353f7cf, %rcx
% Here is the extra copy
movq %rdx, %rax
mulq %rcx
...
    \end{lstlisting}

  % movq %rdx, %rcx
  % shrq $0x4, %rcx
  % imulq $0x3e8, %rcx, %rcx
  % subq %rcx, %rdi
  % imulq $0xf4240, %rdi, %rcx
  % movq %rdx, %rax
  % shrq $0x4, %rax
  % movq %rcx, %rdx
  % retq

% Rewritten
    \begin{lstlisting}
movq %rdi, %rax
shrq $0x3, %rax
movq $0x20c49ba5e353f7cf, %rdx
mulq %rdx
...
    \end{lstlisting}
    \centering
    \caption{Optimized code that beats \texttt{gcc -Og} by eliminating a register copy.}
    \label{fig:elim_copy}

@@

\begin{figure}
    \centering
    % gcc -Og
    \begin{lstlisting}
cmpl %edi, %esi     
movl %edi, %eax    
cmovgel %esi, %eax 
retq 
    \end{lstlisting}
    % Rewrite
    \begin{lstlisting}
cmpl %edi, %esi     
jle .L1
movl %esi, %eax
retq
.L1:
movl %edi,%eax
retq
    \end{lstlisting}
    \caption{Code rewritten to use conditional branches instead of conditional moves by the model.}
    \label{fig:conditional_move}
\end{figure}

@@

\begin{figure}
%-O0
\begin{lstlisting}
pushq %rbp
movq %rsp, %rbp
movb $0x0, -0x1(%rbp)
movb $0x18, -0x2(%rbp)
mozwl -0x2(%rbp), %eax
popq %rbp
retq
\end{lstlisting}
Rewrite
\begin{lstlisting}
movl $0x18, %eax
retq
\end{lstlisting}
    \centering
    \caption{Optimized code that reduces three instructions and three memory operations to one instruction with no memory operations.}
    \label{fig:bitvector}
\end{figure}

@@

\es{Very low priority, but one thing we might want to add for Future Work is to set live out parameters by parsing the C function header instead of our approximation algorithm.}

>>>>

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{booktabs} % for professional tables

\usepackage{color}

@@

\lstdefinelanguage
    [hexrays]{Assembler}
    [x86masm]{Assembler}
    {morekeywords={addl,popq,pushq,movq,movl,retn,rbp,rsp},deletendkeywords={dword,ptr,short},morendkeywords={var1,var2}
    }

@@

%We perform a two-step process for learning program optimization.
In this section we elaborate on our approach that learns to optimize real-world C programs. 
we first describe the underlying neural model before introducing the learning algorithm.

%In this section, we first describe our underlying model, then describe the first step of pre-training the model using imitation learning, and finally describe the second step of using a hill-climbing algorithm that directly minimizes the cost of model-optimized programs \gn{Does it actually minimize the expected cost (in a mathematical sense)? I don't think so.}\pcyin{Done: deleted ``expected''}.
%\pcyin{We need to change the term ``fine-tuning''. Can we say ``RL'' here? Perhaps not because it's not exactly RL?}
%using the outputs of the neural model itself. 
%For the first phase, the dataset of 1.77 million functions mined from open source projects was used. In the second phase we fine-tune the pre-trained model using an iterative-hill climbing algorithm.

\subsection{Neural Program Optimizer}

@@

Our program optimization model $f_{\theta}$ is a neural sequence-to-sequence network, where the input (original) and the output (optimized) programs are represented as sequences of tokens.
Specifically, $f_{\theta}$ is parameterized with a standard Transformer-based encoder \citep{vaswani2017attention}, but with a LSTM-based decoder \citep{hochreiter1997long,chen-etal-2018-best} instead.
This was mainly done to speed-up prediction within the iterative training loop described in \S\ref{sec:hillclimbing}, as transformers suffer from quadratic decoding time in sequence length \citep{chen-etal-2018-best}. We defer more implementation-specific details for our experiments to Section \ref{sec:experiments}.

@@

\subsection{Learning Algorithm}

To learn to optimize complex real-world programs, we develop a two-stage learning approach.
First, in a pre-training step, to capture commonly-used optimization heuristics adopted by existing optimizing compilers, we use supervised learning to train the model on the mined corpus $D_o$ of \texttt{GCC}-optimized programs (E.q. 2) described in \cref{sec:benchmark}. 
Next, to discover more efficient optimization strategies, we propose an  hill-climbing algorithm, where the model directly explores the space of possible optimizations to minimize the cost of predicted optimized programs.

%We use an intuitive iterative hill-climbing algorithm algorithm for learning program optimizations. 
%It consists of a two-step process in which an exploration batch $B_{\mathrm{ex}}$ is sampled from the model. 
\cref{alg:hill_climbing} 
It consists of a two-step process. 
First, an exploration batch $B_{\mathrm{ex}}$ is sampled from the dataset $D$ initialized with mined programs $F^i_u$ and their compiler-optimized outputs $F^i_o$. 
For each input program $F^i_u$ in $B_{\mathrm{ex}}$, we sample a model-predicted optimization $\hat{F}^u_o$, and run $\hat{F}^u_o$ on the test the I/O test suite to compute the cost function $\mathcal{C}(\cdot)$.
%Each output from the batch is executed on the testcases to test equivalence and its performance characteristics are measured and profiled. 
If any of the new samples are both functionally equivalent by our distance function $\mathcal{D}(\cdot)$ and also more optimal compared to the compiler-optimized targets in the original dataset, the compiler-optimized target in the dataset is then replaced with the model's newly-discovered optimal rewrite. 
\pcyin{For each $F^i_u$, are multiple samples of $\hat{F}^i_o$ generated? If that's the case we need to modify Algo 1 to clarify this.}
After the exploration step is taken, a separate training batch $B_{\mathrm{tr}}$ is sampled for maximum-likelihood training from the dataset which may now contain model-optimized targets. \pcyin{The algorithm sketch here is good. We could improve the narrative by connecting each chunks of sentences to line numbers in Algo 1. See \url{https://www.overleaf.com/read/qmxxdyrnqjft}(lines 139 to 164) for an example.}

%\subsection{Pre-training}

%In the pre-training step, we train a neural model learn common optimizations performed by an optimizing compiler (e.g., \texttt{GCC}) using maximum likelihood estimation. \alex{I think we could remove this section, and mention we do pre-training before hill-climbing in section 6}

@@

%\subsection{Iterative Hill-Climbing}

%\pcyin{When explaining a model/algorithm, it's important to first briefly describe the intuition behind each component/sub-section at the very beginning. Can we say something like the follows here: With the model pre-trained on compiler-optimized examples, we then aim to learn more efficient optimization strategies by directly minimizing the expected cost of model-optimized programs.}

%%Commented by Pengcheng: 
Policy gradient algorithms seem like a natural fit for learning how to optimize programs; however, methods based on maximum-likelihood estimation, or imitation learning, provide advantages in terms of stability. Training large neural networks with policy gradient algorithms can be difficult to tune. These difficulties are often attributed to the gradient's high variance which is exacerbated by other issues in program synthesis: a general sparsity of rewards, credit assignment over long sequences, and brittleness of program semantics.

@@

%Commented by Pengcheng: 
If an expert policy or expert examples are available, imitation learning can be desirable, because of its stability. Impressive results have been achieved in the game of computer Go and other games by minimizing the cross-entropy between a policy and an expert-like Monte Carlo tree search policy \cite{silver2017mastering}. In program synthesis from I/O example tasks, supervised learning on example programs has been shown to be competitive, and at times even superior to RL \cite{bunel2018leveraging}. The effectiveness of supervised, or imitation learning, on program synthesis tasks is even more surprising, given the programs used as targets were generated from random generation from a grammar. Given that real-world programs demonstrate a degree of naturalness, and thus, a higher degree of predictability and repetition \cite{hindle2012naturalness}, one may expect imitation learning to generally perform even better in practice on general program synthesis tasks. \alex{I feel these 2 sentences are good points but could be a little too much detail.}\pcyin{I agree. We could try shortening this paragraph once Section 6 is nearly done.}

@@

% Commented by Pengcheng
% Inspired by the success of imitation learning in program synthesis, in this paper we propose an intuitive hill-climbing algorithm, a special case of imitation learning, for learning program optimizations. 
% %We use an intuitive iterative hill-climbing algorithm algorithm for learning program optimizations. 
% %It consists of a two-step process in which an exploration batch $B_{\mathrm{ex}}$ is sampled from the model. 
% It consists of a two-step process. 
% First, an exploration batch $B_{\mathrm{ex}}$ is sampled from the dataset $D$ initialized with mined programs $F^i_u$ and their compiler-optimized outputs $F^i_o$. 
% For each input program $F^i_u$ in $B_{\mathrm{ex}}$, we sample a model-predicted optimization $\hat{F}^u_o$, and run $\hat{F}^u_o$ on the test the I/O test suite to compute the cost function $\mathcal{C}(\cdot)$.
% %Each output from the batch is executed on the testcases to test equivalence and its performance characteristics are measured and profiled. 
% If any of the new samples are both functionally equivalent by our distance function $\mathcal{D}(\cdot)$ and also more optimal compared to the compiler-optimized targets in the original dataset, the compiler-optimized target in the dataset is then replaced with the model's newly-discovered optimal rewrite. 
% \pcyin{For each $F^i_u$, are multiple samples of $\hat{F}^i_o$ generated? If that's the case we need to modify Algo 1 to clarify this.}
% After the exploration step is taken, a separate training batch $B_{\mathrm{tr}}$ is sampled for maximum-likelihood training from the dataset which may now contain model-optimized targets. \pcyin{The algorithm sketch here is good. We could improve the narrative by connecting each chunks of sentences to line numbers in Algo 1. See \url{https://www.overleaf.com/read/qmxxdyrnqjft}(lines 139 to 164) for an example.}

@@

We report our performance on a randomly-sampled 861 examples from the training set as well as the entire test set. Our intent in doing so is to evaluate the learned model in two task-settings: the first being an \textit{inductive} task where we are interested in the model generated and its ability to generalize to unseen examples in the test set. The second task is a \textit{transductive} task where we are not interested in the model per-se, rather, we are interested in the capacity of the model and the training procedure to transduce unoptimized programs to optimized programs. In both tasks we generate samples using a beam-size of 10, executing all on the test suite $\{IO\}_{k}^K$ to calculate both $\mathcal{C}$  and $\mathcal{D}$ and we optionally use the SMT-based verifier to also calculate $D$. In our experiments, whenever we utilize the SMT-based verifier, we perform this in addition to passing all testcases with full correctness. \footnote{Another way to perform transductive evaluation is by reporting statistcs on the best outputs generated during training. We felt beam-search would provide a more interesting comparison.}. Then, of these ten samples, we report results on the best-performing correct sample (if one was available).

@@

We subsequently performed a manual evaluation of the outputs. We randomly sampled 29 of the programs that beat the \texttt{GCC -Og} benchmark in the hold-out set. Upon close analysis, we found only 8 of these 29 programs (or 27.59\%) were semantics preserving and correct. One group of errors was triggered by the model removing conditional branches from programs that were unlikely to be triggered through random testing. 
%
%We witnessed that the model often %removed entire branches from programs. 
%\pcyin{Sorry I don't quite get the %following two sentences.}
Take an example in which we have a program in which we would like to trigger behavior if the variable $x$ is equal to the constant 0 (i.e. $\textbf{if} \;\; \mathbf{x\;==\;0:\; \textsc{do}...}$). If $x$ is represented by a 64-bit integer in a cpu register, and if we use a uniform random distribution to generate values for x in testcases,  likelihood of sampling the exact constant 0 is $(2^{-64})$, and thus very unlikely. As a result, randomly-generated testcases can be highly insufficient for measuring program correctness.

@@

Given the limitations of random testcase generation, we sought to utilize the SMT-based verifier to calculate $D$ for evalutation. When evaluating with SMT solvers, the proportion of programs that outperform the -Og baseline drops dramatically.\footnote{We took an additional step and sampled all 44 examples that outperformed the \texttt{GCC -Og} benchmark in the test set that we could confidently decide on correctness. We found that only 4 were incorrect due to strange edge cases of unlinked assembly and potential complications with the validator.} Nevertheless, the fine-tuning procedure still allows for a noticeable improvement over the pre-trained baseline. Additionally, after the hill-climbing procedure, the model is less capable of generating samples that outperform the \texttt{GCC -O0} baseline. Upon further inspection, for both samples from the training and the test set 100\% of the programs that were not capable of outperforming the -O0 baseline were deemed logically incorrect by the verifier. By performing the hill-climbing training procedure with only testcases to measure correctness $\mathcal{D}$, the model learned new and valid optimizations; yet it evidently lost its ability to consistently perform less-aggressive \alex{"aggressive" is a little awkward here (I wrote it)}, semantics-preserving optimizations.

@@

\alex{Decoding for this is scheduled to finish at 5:30 PM eastern. When I evaluated on the dev-set, it seems that the proportion that ``beat" GCC was below the testcase-trained, but it didn't degrade as much on the -O0 task. I think it is possible the model sampled too few "rewrites" early on and started over-fitting to the -Og examples. Will confirm}

Given that the hill-climbing algorithm, when trained with testcases only for correctness, learned spurious optimization behavior, we sought to perform the hill-climbing algorithm using the validator to calculate $D$ inside the training loop. We then evaluated this model only using both testcases as well as the solver. Our results are located in Table \ref{tab:z3_inner_loop_results}.

Utilizing the solver in the inner-loop of training did not cause a significant degradation in performance in the task of outperforming the -O0 baseline the way training on testcases only did, and on both metrics we see the fine-tuning process enabled the model to outperform the pre-trained baseline model. For thoroughness, we sampled \alex{XX} programs that outperformed the GCC -Og benchmark and manually evaluated their correctness and saw \alex{XX.XX\%, i.e. very few} were incorrect. A full breakdown of the manual analysis results are located in \ref{tab:manual}.

\begin{table}[t]
\caption{Evaluation results for model trained and evaluated on testcases only. The table reports the proportion of programs within the dataset whose best sample through beam search outperformed either the \texttt{GCC -O0} baseline or the \texttt{GCC -Og} baseline}

\label{tab:random_generation_results}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcc}
\toprule
Task & Outperf. -O0 & Outperf. -Og \\
\midrule
Pre-Trained trn   & 89.43\% & 6.04\% \\
Pre-Trained tst    & 90.33\% & 5.67\% \\
Hill-Climbing trn  & 93.14\% & 24.27\% \\
Hill-Climbing tst  & 87.56\% & 22.22\% \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\begin{table}[t]
\caption{Evaluation results for models evaluated with the SMT-based verifier as well as the testcase suite for correctness. The table report the proportion of programs sampled outperforming GCC settings. A prefix of \textsc{TC} indicates the model trained through hill-climbing with testcases only for correctness, and \textsc{SMT} indicated the model trained through hill-climbing with the SMT-based verifier and testcases for correctness}
% \caption{Evaluation results for model fine-tuned on testcases-only and evaluated with testcases and the SMT solver for correcteness. The table report the proportion of programs sampled outperforming GCC settings. The inductive task is denoted by \textit{ind.} and the transductive task is denoted by \textit{trans.}}
\label{tab:z3_results}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{lcc}
\toprule
Task & Outperf. -O0 & Outperf. -Og \\
\midrule
Pre-Trained trn   & 85.93\% & 2.21\% \\
Pre-Trained tst    & 86.22\% & 1.11 \% \\
TC Hill-Climbing trn  & 65.33\% & 5.22\% \\
TC Hill-Climbing tst  & 74.69\% & 5.81\% \\
SMT Hill-Climbing trn  & --.--\% & --.--\% \\
SMT Hill-Climbing tst  & --.--\% & --.--\% \\
\bottomrule
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

% \begin{table}[t]
% \caption{Evaluation results for model fine-tuned and evaluated with both testcases and the SMT solver for determining correctness. The table report the proportion of programs sampled outperforming GCC settings. The inductive task is denoted by \textit{ind.} and the transductive task is denoted by \textit{trans.}}
% \label{tab:z3_inner_loop_results}
% \vskip 0.15in
% \begin{center}
% \begin{small}
% \begin{sc}
% \begin{tabular}{lcc}
% \toprule
% Task & Outperf. -O0 & Outperf. -Og \\
% \midrule
% Pre-Trained ind.    & --.-\% & --.--\% \\
% Hill-Climbing ind.  & --.--\% & --.--\% \\
% Pre-Trained trans.   & --.--\% & --.--\% \\
% Hill-Climbing trans.  & --.--\% & --.--\% \\
% \bottomrule
% \end{tabular}
% \end{sc}
% \end{small}
% \end{center}
% \vskip -0.1in
% \end{table}

% \alex{let's remove this table for now; we've summarized our findings in the body and the footnotes}
% \begin{table}[t]
% \caption{This table reports the proportion of programs that manual evaluation determined correct. This was performed only for the examples within the hold-out test set for the inductive task and the models trained with the iterative hill-climbing algorithm and for the programs that outperformed GCC -Og.  \textbf{TC} corresponds to experiments where correctness was determined only by passing all testcases; whereas, \textbf{SMT} corresponds to experiments where correctness was determined by passing all testcases \textit{and} verifying via the SMT solver}
% \label{tab:manual}
% \vskip 0.15in
% \begin{center}
% \begin{small}
% \begin{sc}
% \begin{tabular}{lc}
% \toprule
% Task & Proportion Verified \\
% \midrule
% TC Train/TC Eval  & 22.72\%\\
% TC Train/Z3 Eval   & 91.30\%\\
% Z3 Train/Z3 Eval  & --.--\%\\
% \bottomrule
% \end{tabular}
% \end{sc}
% \end{small}
% \end{center}
% \vskip -0.1in
% \end{table}

@@

\newcommand{\ozerocodecaption}{-O0 code\xspace}
\newcommand{\ogcodecaption}{-Og code\xspace}
\newcommand{\modelcodecaption}{Model-optimized code\xspace}
\newcommand{\codeboxwidth}{.47\columnwidth}

@@

Most x86-64 instructions allow operands to be stored in any register.  A notable exception is the \ed{decorate me} \texttt{mulq} instruction, which multiplies two 64-bit operands and stores the 128-bit result without truncation.
%
Though this instruction multiplies two numbers, it only takes one operand as an argument.  By definition, \texttt{mulq op} computes the multiplication of \texttt{\%rax} and \texttt{op}, and then stores the upper 64-bits of the result in \%rdx and the lower bits in \%rax.  The registers \%rax and \%rdx are fixed and cannot be changed.
%
\ed{@pcyin Can we put the two listings side by side?  We can truncate the long constant, it's not important.  It would also be nice if we could highlight the changes.}
\begin{figure}
% -Og compiler code
%\subcaptionbox{ok}{there}
%\subfigure[ok]{

    \begin{subfigure}[t]{\codeboxwidth}
    \begin{lstlisting}[language={[x64]Assembler},escapechar=!] 
movq %rdi, %rdx
shrq $0x3, %rdx
movq $0x20..., %rcx
!\colorbox{red}{movq \%rdx}!
% Here is the extra copy
\colorbox{red}{movq %rdx, %rax}
mulq %rcx
...
    \end{lstlisting}
    \caption{\ogcodecaption}
    \end{subfigure}
    \hfil
  % movq %rdx, %rcx
  % shrq $0x4, %rcx
  % imulq $0x3e8, %rcx, %rcx
  % subq %rcx, %rdi
  % imulq $0xf4240, %rdi, %rcx
  % movq %rdx, %rax
  % shrq $0x4, %rax
  % movq %rcx, %rdx
  % retq
%
% Rewritten
    \begin{subfigure}[t]{\codeboxwidth}
    \begin{lstlisting}
movq %rdi, %rax
shrq $0x3, %rax
movq $0x20..., %rdx
mulq %rdx
...
    \end{lstlisting}
    \caption{\modelcodecaption}
    \end{subfigure}
    \centering
    \caption{Optimized code that beats \texttt{gcc -Og} by eliminating a register copy.}
    \label{fig:elim_copy}

@@

\begin{figure}
    \centering
    % gcc -Og
    \begin{subfigure}[t]{.45\columnwidth}
    \begin{lstlisting}[language={[x64]Assembler}]
cmpl %edi, %esi     
movl %edi, %eax    
cmovgel %esi, %eax 
retq 
    \end{lstlisting}
    \caption{\ogcodecaption}
    \end{subfigure}
    \hfil
    % Rewrite
    \begin{subfigure}[t]{.45\columnwidth}
    \begin{lstlisting}
cmpl %edi, %esi     
jle .L1
movl %esi, %eax
retq
.L1:
movl %edi,%eax
retq
    \end{lstlisting}
    \caption{\modelcodecaption}
    \end{subfigure}
    \caption{Code rewritten to use conditional branches instead of conditional moves by the model.}
    \label{fig:conditional_move}
\end{figure}

@@

\begin{figure}
%-O0
\begin{subfigure}[t]{\codeboxwidth}
\begin{lstlisting}
pushq %rbp
movq %rsp, %rbp
movb $0x0, -0x1(%rbp)
movb $0x18, -0x2(%rbp)
mozwl -0x2(%rbp), %eax
popq %rbp
retq
\end{lstlisting}
\caption{\ozerocodecaption}
\end{subfigure}
% Rewrite
%\hfil
\begin{subfigure}[t]{\codeboxwidth}
\begin{lstlisting}
movl $0x18, %eax
retq
\end{lstlisting}
\caption{\modelcodecaption}
\end{subfigure}
    \centering
    \caption{Optimized code that reduces three instructions and three memory operations to one instruction with no memory operations.}
    \label{fig:bitvector}
\end{figure}

@@

\ed{Very low priority, but one thing we might want to add for Future Work is to set live out parameters by parsing the C function header instead of our approximation algorithm.}
